Programming Language Foundations in Agda

### 2019-07-22 papers/scp/PLFA.tex

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can I find it in the literature.  Third, that
  using extrinsically-typed terms is far less
  perspicuous than using intrinsically-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

% NEW
Some researchers are clearly familiar with the connection between
progress and preservation and animation.  In private correspondence,
Bob Harper referred to it as the \emph{pas de deux}, a dance between
progress, which takes well-typing to a step, and preservation, which
takes a step back to well-typing.  Nonetheless, neither the technique
nor the appealing terminology appears in \citet{Harper-2016}.  The
appeal to the Agda mailing list bore late fruit: Oleg Kiselyov
directed me to unpublished remarks on his web page where he uses the
name \texttt{eval} for a proof of progress and notes ``the very proof
of type soundness can be used to evaluate sample expressions''
\citep{Kiselyov-2009}.  Nonetheless, as of this writing, we still have
not located a mention in the published literature.

% NEW
There are places in the literature where one might expect a remark on
the relation between progress and preservation and animation---but no
such remark appears.  In the PoplMark Challenge
\citep{Aydemir-et-al-2005}, Challenge~2A is to prove progress and
preservation for System F$_{<:}$, while Challenge~3 is to prove
animation for the same system. Nowhere do the authors indicate that in
an intuitionistic logic these are essentially the same problem.
\cite{Owens-et-al-2016}, when discussing extraction of animators for
small-step semantics, mention Redex and K, but no other possibilities.
We hope the stress in PLFA on the fact that in an intuitionistic
setting progress and preservation imply animation will mean that the
connection becomes more widely known.

@@

Philip now has five years of experience teaching from SF and one year
teaching from PLFA.  To date, he has taught three courses from PLFA.
\begin{itemize}
\item
  University of Edinburgh, September--December 2018 (with teaching
  assistance from Wen and Chad Nester); twenty 2-hour slots,
  comprising one hour of lecture followed by one hour of lab. Ten
  students completed the course, fourth-year undergraduates and
  masters.  The course covered Parts~I and~II of PLFA.

\item
  Pontifícia Universidade Católica do Rio de Janeiro (PUC-Rio),
  March--July 2019, hosted by Roberto Ieuramalischy; ten 3-hour
  slots, comprising two hours of lecture followed by one hour of lab.
  Ten students completed the course, mostly doctoral students.  The
  course covered Parts~I and~II of PLFA, save students read chapter
  Lists on their own, and chapter Bisimilarity was skipped.

>>>>

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can we find it in the literature.  Third, that
  using extrinsically-typed terms is far less
  perspicuous than using intrinsically-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

% NEW
Some researchers are clearly familiar with the connection between
progress and preservation and animation.  In private correspondence,
Bob Harper referred to it as the \emph{pas de deux}, a dance between
progress, which takes well-typing to a step, and preservation, which
takes a step back to well-typing---but neither the technique
nor the appealing terminology appears in \citet{Harper-2016}.  The
appeal to the Agda mailing list bore late fruit: Oleg Kiselyov
directed us to unpublished remarks on his web page where he uses the
name \texttt{eval} for a proof of progress and notes ``the very proof
of type soundness can be used to evaluate sample expressions''
\citep{Kiselyov-2009}.  Nonetheless, as of this writing, we still have
not located a mention in the published literature.

% NEW
There are places in the literature where one might expect a remark on
the relation between progress and preservation and animation---but no
such remark appears.  In the PoplMark Challenge
\citep{Aydemir-et-al-2005}, Challenge~2A is to prove progress and
preservation for System~F$_{<:}$, while Challenge~3 is to prove
animation for the same system. Nowhere do the authors indicate that in
an intuitionistic logic these are essentially the same problem.
\cite{Owens-et-al-2016}, when discussing extraction of animators for
small-step semantics, mention Redex and K, but no other possibilities.
We hope the stress in PLFA on the fact that in an intuitionistic
setting progress and preservation imply animation will mean that the
connection becomes more widely known.

@@

Philip now has five years of experience teaching from SF and one year
teaching from PLFA.  To date, he has taught three courses from PLFA.
\begin{itemize}
\item
  University of Edinburgh, September--December 2018 (with teaching
  assistance from Wen and Chad Nester); twenty 2-hour slots,
  comprising one hour of lecture followed by one hour of lab. Ten
  students completed the course, fourth-year undergraduates and
  masters.  The course covered Parts~I and~II of PLFA, up through
  chapter Untyped.

\item
  Pontifícia Universidade Católica do Rio de Janeiro (PUC-Rio),
  March--July 2019, hosted by Roberto Ieuramalischy; ten 3-hour
  slots, comprising two hours of lecture followed by one hour of lab.
  Ten students completed the course, mostly doctoral students.  The
  course covered Parts~I and~II of PLFA, up through chapter Untyped,
  save students read chapter
  Lists on their own, and chapter Bisimilarity was skipped.
